---
title: "Survival Outcome of Allogeneic Hematopoietic Stem Cell Transplantation Patients"
output: html_notebook
---

# 1. Import Libraries
```{r}
# Load required packages from the tidyverse and additional libraries
library(tidyverse)       # Data manipulation and visualization
library(caret)           # Data partitioning and model evaluation
library(gridExtra)       # Plot arrangement
library(Boruta)          # Feature selection
library(Epi)             # ROC optimal threshold determination
library(ggcorrplot)      # Correlation visualization
library(dplyr)           # Data manipulation (subset of tidyverse)
library(haven)           # Importing SAS datasets
library(GGally)          # Pair plots
library(rsample)         # Resampling, splitting data
library(pROC)            # ROC analysis
library(car)             # Variance inflation factor (VIF)
library(reshape2)        # Data reshaping for plots
```
# 2. Load and Explore the Dataset
```{r}
# Read the SAS dataset
df <- read_sas("RT14-02.sas7bdat")

# Quick look at the data
head(df)
str(df)
summary(df)

# Check for missing values
colSums(is.na(df))

```
# 3. Data Wrangling
We first remove a few columns that are not needed for analysis, then replace placeholder values (e.g., 99 or 8/9 for gender) with NA before converting categorical variables to factors.
```{r}
# Remove unnecessary columns
data <- df %>% select(-intxtma, -intxrenfail, -intxagvhd24, -intxagvhd34)

# Verify missing values after column removal
colSums(is.na(data))

# Replace placeholder '99' (or '8/9' in 'sex') with NA for selected variables
data$dead[data$dead == 99] <- NA
data$karnofcat[data$karnofcat == 99] <- NA
data$hctcigp[data$hctcigp == 99] <- NA
data$death_gp[data$death_gp == 99] <- NA
data$orgfail[data$orgfail == 99] <- NA
data$sex[data$sex %in% c(8, 9)] <- NA
data$racegp[data$racegp == 99] <- NA
data$kidfunc[data$kidfunc == 99] <- NA
data$drcmvgp[data$drcmvgp == 99] <- NA
data$dnrtype[data$dnrtype == 99] <- NA
data$priautogp[data$priautogp == 99] <- NA
data$condgp[data$condgp == 99] <- NA
data$atgcampathgp[data$atgcampathgp == 99] <- NA
data$tma_event[data$tma_event == 99] <- NA
data$plsmpshi[data$plsmpshi == 99] <- NA
data$agvhdgp[data$agvhdgp == 99] <- NA
data$agvhd24[data$agvhd24 == 99] <- NA
data$agvhd34[data$agvhd34 == 99] <- NA
data$dwoagvhd24[data$dwoagvhd24 == 99] <- NA
data$dwoagvhd34[data$dwoagvhd34 == 99] <- NA
data$renfail[data$renfail == 99] <- NA
data$dworenfail[data$dworenfail == 99] <- NA

# Confirm the missing value counts after replacement
colSums(is.na(data))

```
## Converting Variables to Factors
The following code transforms key variables into factors with meaningful labels.
```{r}
data$dead <- factor(data$dead, levels = c(0, 1), labels = c("Alive", "Dead"))
data$karnofcat <- factor(data$karnofcat, levels = c(1, 2), labels = c("90-100", "<90"))
data$hctcigp <- factor(data$hctcigp, levels = c(0, 1, 2, 3),
                       labels = c("No comorbidity", "1", "2", ">=3"))
data$death_gp <- factor(data$death_gp,
                        levels = 0:9,
                        labels = c("Alive", "Primary Disease", "Graft rejection/failure",
                                   "GVHD", "Infection", "Organ failure", "Second malignancy",
                                   "Hemorrhage", "Vascular", "Other"))
data$sex <- factor(data$sex, levels = c(1, 2), labels = c("Male", "Female"))
data$racegp <- factor(data$racegp, levels = c(1, 2, 3),
                      labels = c("Caucasian", "African-American", "Others"))
data$kidfunc <- factor(data$kidfunc, levels = c(0, 1),
                       labels = c("Decreased kidney function", "Normal kidney function"))
data$drcmvgp <- factor(data$drcmvgp, levels = c(1, 2, 3, 4),
                       labels = c("+/+", "+/-", "-/+", "-/-"))
data$dnrtype <- factor(data$dnrtype,
                       levels = c(0,1,2,3,4,5,6,7,8,9,11,12,13,14),
                       labels = c("Autologous", "HLA-identical sibling", "Twin",
                                  "Other related", "Well-matched unrelated",
                                  "Partially-matched unrelated", "Mis-matched unrelated",
                                  "Multi-donor", "Unrelated, HLA match unknown", "Cord blood",
                                  "CB 6/6", "CB 5/6", "CB <=4/6", "CB, HLA match unknown"))
data$priautogp <- factor(data$priautogp, levels = c(0, 1), labels = c("No", "Yes"))
data$condgp <- factor(data$condgp, levels = c(1, 2, 3, 4),
                      labels = c("Myeloablative TBI", "Myeloablative Bu based", "Other myeloablative", "RIC/NMA"))
data$atgcampathgp <- factor(data$atgcampathgp, levels = c(1, 2, 3, 4),
                            labels = c("ATG + CAMPATH", "ATG alone", "CAMPATH alone", "No ATG or CAMPATH"))
data$tma_event <- factor(data$tma_event, levels = c(0, 1), labels = c("No", "Yes"))
data$plsmpshi <- factor(data$plsmpshi, levels = c(0, 1), labels = c("No", "Yes"))
data$agvhdgp <- factor(data$agvhdgp, levels = c(0, 1, 2),
                       labels = c("No aGVHD or grade 1", "Grade 2", "Grade 3-4"))
data$agvhd24 <- factor(data$agvhd24, levels = c(0, 1), labels = c("No", "Yes"))
data$agvhd34 <- factor(data$agvhd34, levels = c(0, 1), labels = c("No", "Yes"))
data$dwoagvhd24 <- factor(data$dwoagvhd24, levels = c(0, 1), labels = c("No", "Yes"))
data$dwoagvhd34 <- factor(data$dwoagvhd34, levels = c(0, 1), labels = c("No", "Yes"))
data$renfail <- factor(data$renfail, levels = c(0, 1), labels = c("No", "Yes"))
data$dworenfail <- factor(data$dworenfail, levels = c(0, 1), labels = c("No", "Yes"))

# Confirm factor conversions
str(data)

```
# 4. Data Preparation for Modeling
We remove the variable at column 27 (if not needed), omit rows with missing values, and check the outcome distribution.
```{r}
# Remove column 27 & 31 and drop any rows with NAs
working_df <- data[,c(-27,-31)]
working_df <- na.omit(working_df)

# Check column names and structure
colnames(working_df)
str(working_df)

# Outcome distribution and demographics
table(working_df$dead)
table(working_df$sex)

```
# 5. Exploratory Data Analysis (EDA)
Visualize distributions of key variables and their relationship to survival status.
```{r}
# Age distribution
ggplot(working_df, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "lightblue", color = "black") +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()

# Proportion of survival outcome by sex
ggplot(working_df, aes(x = sex, fill = dead)) +
  geom_bar(position = "fill") +
  labs(title = "Outcome Proportion by Sex", x = "Sex", y = "Proportion") +
  theme_minimal()

# Overall outcome distribution
ggplot(working_df, aes(x = dead)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Overall Survival Outcome", x = "Outcome", y = "Count") +
  theme_minimal()

```
```{r}
# Age distribution by outcome
ggplot(working_df, aes(x = age, fill = dead)) +
  geom_histogram(position = "dodge", binwidth = 5, color = "black") +
  labs(title = "Age Distribution by Outcome", x = "Age", y = "Count") +
  theme_minimal()

# Race group by outcome
ggplot(working_df, aes(x = racegp, fill = dead)) +
  geom_bar(position = "fill") +
  labs(title = "Outcome Proportion by Race Group", x = "Race Group", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal()

# Time from HCT to death/last follow-up
ggplot(working_df, aes(x = intxsurv)) +
  geom_histogram(fill = "lightgreen", color = "black", bins = 30) +
  labs(title = "Distribution of Time from HCT to Follow-Up",
       x = "Time from HCT", y = "Count") +
  theme_minimal()

#Correlation/Pair Plot among Continuous Features
numeric_vars <- working_df %>% select(where(is.numeric))

ggpairs(
  numeric_vars,
  upper = list(continuous = wrap("cor", size = 3)),  # Keep correlation in upper
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8)),  # Transparent, smaller points
  diag  = list(continuous = wrap("densityDiag", alpha = 0.6))  # Smooth KDE on diagonal
) +
  ggtitle("Pairwise Relationships Among Continuous Features") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 10),
    panel.grid = element_blank()
  )
```
# 6. Data Splitting and Feature Selection

Split the dataset into training (70%) and testing (30%) sets and use the Boruta algorithm for feature selection.

```{r}
set.seed(123)  # For reproducibility
split <- initial_split(working_df, prop = 0.7)
training <- training(split)
testing  <- testing(split)

# Perform feature selection with Boruta
boruta.analysis <- Boruta(dead ~ ., data = training)
print(boruta.analysis)
print(boruta.analysis$finalDecision)

```
# 7. Logistic Regression Modeling

Fit a logistic regression model using selected predictors. Adjust the model formula based on domain expertise and Boruta results.
```{r}
model <- glm(dead ~ age + racegp + karnofcat + hctcigp + 
               diseasegp + atgcampathgp + graftgp + drabomatch + 
               kidfunc + condgp + malig_grp + yeartx + drcmvgp +
               priautogp + plsmpshi + intxsurv + renfail + dnrtype + agvhdgp,
             data = training,
             family = binomial)


```
# 8. Model Evaluation

Evaluate the model on the testing set with prediction probabilities, confusion matrix, ROC curve, and assess multicollinearity with VIF.
```{r}
# Predict probabilities on testing data
testing$predicted_prob <- predict(model, newdata = testing, type = "response")

# Convert probabilities to binary predictions using a cutoff of 0.5
testing$predicted_class <- ifelse(testing$predicted_prob >= 0.5, "Dead", "Alive")
testing$predicted_class <- factor(testing$predicted_class, levels = c("Alive", "Dead"))

# Confusion Matrix
conf_matrix <- confusionMatrix(testing$predicted_class, testing$dead)
print(conf_matrix)

# ROC Curve and AUC
roc_curve <- roc(testing$dead, testing$predicted_prob)
plot(roc_curve, col = "blue", main = "ROC Curve for Logistic Regression Model")
auc_value <- auc(roc_curve)
print(auc_value)

# Evaluate multicollinearity with VIF
vif_values <- vif(model)
print(vif_values)

```
# 9. Model Refinement and Risk Stratification
Use stepwise selection for model refinement and stratify patients into risk groups based on predicted probabilities.
```{r}
# Stepwise selection to refine the model
step_model <- step(model, direction = "both")
summary(step_model)

# Calculate risk score and categorize risk groups
testing$risk_score <- predict(model, newdata = testing, type = "response")
testing$risk_category <- cut(testing$risk_score, breaks = c(0, 0.3, 0.6, 1),
                             labels = c("Low Risk", "Medium Risk", "High Risk"))
table(testing$risk_category)

```
# 10. Interpretation of results
```{r}
# Model coefficients as odds ratios with 95% confidence intervals
odds_ratios <- exp(coef(model))
conf_int <- exp(confint(model))
results <- data.frame(Estimate = odds_ratios, 
                      CI_Lower = conf_int[,1], 
                      CI_Upper = conf_int[,2])
print(results)
pct_change <- ifelse(odds_ratios >= 1,
                     (odds_ratios - 1) * 100,      # percent increase
                     (1 - odds_ratios) * 100)      # percent decrease
output <- data.frame(
  Predictor   = names(odds_ratios),
  OR          = round(odds_ratios, 3),
  CI_Lower    = round(conf_int[,1], 3),
  CI_Upper    = round(conf_int[,2], 3),
  Pct_Change  = round(pct_change, 1)  # one decimal place
)
print(output)


```

# 11. Visualization of Key Relationships
Generate additional plots to illustrate relationships between variables and outcomes
```{r}
# Bar plot: Renal failure and survival status
ggplot(training, aes(x = renfail, fill = dead)) +
  geom_bar(position = "fill") +
  labs(title = "Survival Rates by Renal Failure",
       x = "Renal Failure", y = "Proportion") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal()

# Bar plot: GVHD severity and survival status
ggplot(training, aes(x = agvhdgp, fill = dead)) +
  geom_bar(position = "fill") +
  labs(title = "Survival Rates by GVHD Severity",
       x = "GVHD Severity", y = "Proportion") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal()

# Density plot: Age distribution by survival and renal failure
ggplot(training, aes(x = age, fill = dead)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ renfail) +
  labs(title = "Age Distribution by Survival & Renal Failure",
       x = "Age", y = "Density") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal()

# Survival probability over transplant year
ggplot(training, aes(x = yeartx, y = as.numeric(dead == "Alive"))) +
  geom_smooth(method = "loess", color = "blue", fill = "lightblue") +
  labs(title = "Survival Probability Over Transplant Years",
       x = "Transplant Year", y = "Probability of Survival") +
  theme_minimal()

# Bar plot: HCT-CI and GVHD severity with survival
ggplot(training, aes(x = hctcigp, fill = dead)) +
  geom_bar(position = "fill") +
  facet_wrap(~ agvhdgp) +
  labs(title = "Survival by HCT-CI Score & GVHD Severity",
       x = "HCT-CI Score (Comorbidities)", y = "Proportion") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal()

# Violin plot: Age distribution by survival status
ggplot(training, aes(x = dead, y = age, fill = dead)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.2, color = "black", alpha = 0.7) +
  labs(title = "Age Distribution: Survivors vs. Non-Survivors",
       x = "Survival Status", y = "Age") +
  scale_fill_manual(values = c("blue", "red")) +
  theme_minimal()

```
# 12. Assessing Feature Correlations & Variable Effects

Plot a heatmap of the correlation matrix among numeric features and a forest plot to visualize predictor effects.
```{r}
# Correlation heatmap for numeric features
cor_matrix <- cor(training %>% select(where(is.numeric)), use = "complete.obs")
melted_cor <- melt(cor_matrix)
ggplot(melted_cor, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(title = "Heatmap of Feature Correlations", x = "", y = "") +
  theme_minimal()

# Forest plot: Odds ratios of predictors
# Extract coefficients and confidence intervals, then remove intercept
coef_df <- data.frame(
  Predictor = names(coef(model)),
  LogOdds = coef(model)
) %>%
  filter(Predictor != "(Intercept)") %>%
  arrange(LogOdds)

ggplot(coef_df, aes(x = reorder(Predictor, LogOdds), y = LogOdds, fill = LogOdds > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("red", "blue")) +
  labs(title = "Effect Size of Predictors (Log-Odds)",
       x = "Predictor", y = "Log-Odds") +
  theme_minimal()

# Forest plot with Odds Ratios and 95% CI
odds_ratios <- exp(coef(model))
conf_intervals <- exp(confint(model))

forest_df <- data.frame(
  Predictor = names(odds_ratios),
  OR = odds_ratios,
  Lower_CI = conf_intervals[, 1],
  Upper_CI = conf_intervals[, 2]
) %>%
  filter(Predictor != "(Intercept)") %>%
  arrange(OR) %>%
  mutate(
    # Mark significance if CI does not cross 1
    Significant = ifelse(Lower_CI > 1 | Upper_CI < 1, "Yes", "No")
  )

# Plot
ggplot(forest_df, aes(x = OR, y = fct_reorder(Predictor, OR))) +
  geom_point(aes(color = Significant), size = 3) +
  geom_errorbarh(aes(xmin = Lower_CI, xmax = Upper_CI, color = Significant), height = 0.3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  scale_x_log10() +  # Log scale for better spacing
  scale_color_manual(values = c("Yes" = "darkblue", "No" = "gray")) +
  labs(
    title = "Forest Plot: Odds Ratios for Survival",
    x = "Odds Ratio (log scale)",
    y = "Predictor",
    color = "Significant"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 9),
    plot.title = element_text(hjust = 0.5)
  )

```

